---
title: "lab"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Types

Vector, matrices, data frames.

### How to create a vector

Let's say you want to create a vector `x = [2, 10, 11, 1, 23]`. The easiest way to create this vector is through the `c` function:
```{r, include=TRUE}
x <- c(2, 10, 11, 1, 23)
```

In case you want to know what's the current value of variable x, you can either just add the command `x`. Alternative, you can use `print(x)`:
```{r, include=TRUE}
x # this will print the current value of x

print(x) # this will also print it

```

Now, let's say you want to create a vector with a sequence of numbers, such as creating a vector with the numbers from 1 to 10. There are different ways of accomplishing this, such as:
```{r, include=TRUE}
x <- 1:10
x

```

In this case `1:10` denotes a way of creating a sequence of numbers from 1 to 10. You can also specify it in decreasing order, in case you want to create a sequence from 10 to 1:
```{r, include=TRUE}
x <- 10:1
x

```

You can add to vectors together with the `c()` function:
```{r, include=TRUE}
x <- c(1:5, 5:1)
x

```

```{r}
x <- c(10, 100, 1000) 
y <- c(1, 2, 3) 
z <- c(x, y, x)
z

```

You can also use a different function to create this vector, with the function `seq`:
```{r, include=TRUE}
x <- seq(1:10)
x

```

Alternativity, you can use different syntax:
```{r, include=TRUE}
x <- seq(from=1, to=10)
x

```


So this line `x <- seq(1:10)` is equivalent to `x <- seq(from=1, to=10)`. The latter has some advantages, such as defining other parameters. Let's say that in the sequence you want to have numbers increasing by 2:
```{r, include=TRUE}
x <- seq(from=1, to=10, by=2)
x

```

Or, you want exactly 7 numbers:
```{r, include=TRUE}
x <- seq(from=1, to=10, length.out=7)
x

```

R will automatically create the equally spaced sequence in the [1, 10] bounds. To see more parameters of seq function, check the R help (`?seq`).

There are other way of specifying vectors, with the funcion `rep`. Let's say you want to create a vector of length 10, all of them being 1:
```{r, include=TRUE}
x <- rep(1, 10)
x

```

There are other types of sequences you can do with rep, such as:
```{r, include=TRUE}
rep(1:3)


rep(1:3, times=3)


rep(1:3, each=3)

```

Let's say you want to create a sequence of square numbers, `1  4  9 16 25`. You can do so by creating a vector, and then squaring it. R works as element base, so the square function will be applied to each element of the sequence:
```{r}
x <- 1:5
x


y <- x ^ 2 # alternative, you can use x ** 2
y
## 1  4  9 16 25
```

Now, let's say you want to create the sequence 1, 2, 4, 8, 16, 32. Notice this sequence is 2^n:
```{r}
x <- 2**(0:10)
x 
# [1]    1    2    4    8   16   32   64  128  256  512 1024
```

In this case, a vector is applied as a pow to a single number, creating a sequence of numbers. Note the use of the parenthesis.

### Operations with vectors:

You can do all kind of operations with vectors, +, -, /, etc. Operations are performed element-wise.
```{r}
x <- c(2, 3, 4, 5) 
y <- c(1, 0, 3, 2)

x + y


(x + y) * 2


x + y * 2

```
Note the different values in the third operation, as the `* 2` is applied to y before the sum to x.

### How to access a vector element. Elements in R are index based, with the first index being 1 (instead of 0 like in other programming languages)
```{r}
x <- c(2, 4, 5, 8, 10)
```

In this case, 2 is index 1, 4 is index 2, 5 is index 3.
```{r}
x[1]


x[2]


x[3]

```

If you want to access more than 1 element, you can specify a vector of indexes:
```{r}
x[1:3]

```

However, if you specify indexes that don't exist in the vector, R will return NAs:
```{r}
x[3:10]

```

You can also subset this vector by removing some indices:
```{r}
x

x[-5]

x[-c(4, 5)]
```
The commands above remove index 5 (second command) and 4,5 indices (third command).

You can assign a value from the same way. Let's say you want to change the first number to be 9 instead of 2:
```{r}
x[1]

x[1] <- 9 

x[1]

x
```

You can also assign a sequence of numbers in the same operation. Let's say you want the first thre elements to be 1, 2, 3:
```{r}
x

x[1:3] <- c(1, 2, 3)
x
```

Now let's assume you want to subset the vector based on some logic. If you have a vector from 1:5, you can specify which indices you want by a logic vector:
```{r}
x <- 1:5
x

x[c(TRUE, TRUE, FALSE, FALSE, TRUE)]

# Equivalent to:
x[c(1, 2, 5)]
```

This allow us to perform some logic for subsetting. Let's first see what happens when you perform comparison equations over a vector:
```{r}
x <- 1:5
y <- x < 3
y
```

If we have a vector of 1:5 (elements 1, 2, 3, 4, 5) and we use a comparison method of `x < 3` we obtain a vector with `c(TRUE, TRUE, FALSE, FALSE, FALSE)`. This means that the first 2 elements of vector `x` are `<3` and the rest elements of vector are not. Now we can use that vector to subset x in case we only want those elements:
```{r}
x[y]
```

What if we wanted the oposite? You want the numbers that are equal or higher than 3. You can imagine that an easy way to do it is:
```{r}
x <- 1:5
z <- x >= 3
z

x[z]
```

However, you can use our previous vector and negate it:
```{r}
!y

x[!y]
```

Subsetting though does not require to specify a different variable, you can apply the logic directly:
```{r}
x[x<3]

x[!(x<3)]

x[x>=3]
```

You can use more than one logic operations:
```{r}
x[x<3 | x==5]
```

What if you only want even numbers?
```{r}
x[x%%2 == 0]
```

### Operations with vectors
You can do operations with vectors:
```{r}
x <- c(2, 2, 4, 6, 6)

mean(x)
sum(x)
var(x)
sd(x)
```

Be careful if you want to calculate the mean of the vector and the vector has NA:
```{r}
x <- c(2, 2, 4, 6, 6, NA)
mean(x)
```

In order to calculate the mean of a vector with NA you can do either of the following commands:
```{r}
mean(na.omit(x))
mean(x, na.rm = TRUE)
```
### How to create a matrix


Internally, R stores matrices in column-major mode, i.e. the matrix is stored as [1 2 3 4 5 6 7 8 9] unlike C or Java where it would be stored in what is called "row-major mode", i.e. the rows would be stacked on top of each other.


matrix(c(1, 2, 3, 4, 5, 6), ncol=2)
matrix(NA, ncol=2, nrow=4)

matrix(1, ncol=2, nrow=4)

cbind(c(1,2,3), c(4,5,6))

rbind(
  c(1,2,3), 
  c(4,5,6)
)

M <- cbind(
  c(1,2,3), 
  c(4,5,6)
)

Transpose of a matrix:
t(M)

Inverse of a matrix:
solve(M)

diag(c(1,2,3,4))
